---
title: "Assign8_Choi_GLM"
author: "Serena Choi"
date: "3/21/2022"
output: pdf_document
---
## Setting up

```{r}
duncan <- read.csv("duncan.csv")
colnames(duncan)<-c("occtitle", "occtype", "income", "educ", "prestige")
attach(duncan)
summary(duncan)
library(nnet)
library(dplyr)
table(occtype)
help(multinom)
library(car)
library(carData)
library(MASS)
```

## Using the Duncan data, estimate the effect of prestige scores on occupational type using either a multinomial logistic regression OR an ordered logit model. 

### Ordered logit model
```{r}
help(polr)
ordered <- polr (as.ordered(occtype) ~ prestige  , method="logistic")
summary(ordered)
```

According to the ordered logit model above, for a one-unit change in prestige score, the logit of observing higher level is expected to change by 0.038. 


### Multinomial logistic regression. 
```{r}
multinomial <- multinom(factor(occtype) ~ prestige , data=duncan, method="logistic")
summary(multinomial)
```

According to the mutlinomial logistic model, the effect of the prestige on logit of being in the professional category relative to the blue collar category is -9.319. The effect of the prestige on logit of being in the white-collar category relative to the blue collar category is -2.553. 

##Interpret the effect of prestige scores on occupational type using odds  (in one sentence).   

### ordered 
```{r}
or1 <- exp(coef(ordered))
or1
```

For a one-unit change in prestige, the effect of one-unit change in prestige on the odd of being in a higher category is 1.039.

### multinomial
```{r}
or2 <- exp(coef(multinomial))
or2
```

The coefficients above reflect the odds of being in a given category relative to the blue-collar category. 

## What is the predicted probability of being a professional for those with a prestige score of 10? What about for those with a prestige score of 90?

 I left the sections below as a comment as I kept getting non-conformable arguments below. 
```{r}
###ordered
#beta1 <- coef(ordered) 
#tau1 <- ordered$zeta  # store the intercepts in an object called tau
#X<-cbind(seq(from = 10, to = 90, by = 10)) # looking at the prestige score from 10 to 90
#logit.prob <- function(eta){exp(eta)/(1+exp(eta))}  #the equation for calculating predicted probabilities
#p1 <- logit.prob(tau[1] - X %*% beta)  # calculate the pred probability for blue collar (or category 1)
#p2 <- logit.prob(tau[2] - X %*% beta) - logit.prob(tau[1] - X %*% beta) # calculate the pred probability for prof (or category 2)
#p3 <- 1.0 - logit.prob(tau[2] - X %*% beta) # calculate the pred prob for white collar (or category 3)

#p1	#blue collar
#p2	#professional
#p3 	#white collar
```


```{r}
### multinomial
#beta <- coef(multinomial) # store the coefficients in an object called beta
#tau <- multinomial$zeta  # store the intercepts in an object called tau
#X<-cbind(seq(from = 10, to = 90, by = 10)) # looking at the prestige score from 10 to 90
#logit.prob <- function(eta){exp(eta)/(1+exp(eta))}  #the equation for calculating predicted probabilities
#p1 <- logit.prob(tau[1] - X %*% beta)  # calculate the pred probability for blue collar (or category 1)
#p2 <- logit.prob(tau[2] - X %*% beta) - logit.prob(tau[1] - X %*% beta) # calculate the pred probability for prof (or category 2)
#p3 <- 1.0 - logit.prob(tau[2] - X %*% beta) # calculate the pred prob for white collar (or category 3)

#p1	#blue collar
#p2	#professional
#p3 	#white collar
```


##Using the data on interlocking directorates, estimate the effect of assets and country on the number of interlocking directorates among firms.  

```{r}
ornstein <- read.csv("ornstein.csv")
hist(ornstein$number)	#note the Poisson distribution
table(ornstein$nation)
model3 <- glm(number ~ assets + nation, family=poisson (link ="log"), data=ornstein)
summary(model3)
exp(model3$coef)
```

For every one million dollar increase in assets, the expected count of interlocking directorates increase by 1.00, net of country.

