---
title: "Assign4_Choi_Nonlinear_effects"
author: "Serena Choi"
date: "2/16/2022"
output: pdf_document
---


## Research Question

In this assignment, we will examine the effect of work hours and age on earnings. In examining potential associations, we will focus on whether non-linear effects of independent variables on the response variable exists. The research questions are: 

1) Are there non-linear effects of work hours on earnings? 
2) Are there non-linear effects of age on earnings? 


```{r} 
cps <- read.csv(file = "cps.csv")  #setting the path, data and variable names 
hours <- cps$hours
logwg <- cps$logwg
age <- cps$age
race <- cps$race
educ <- cps$educ
```

## 1. Effect of work hour 

Before delving into the research question, we will first examine the histogram of work hours below. From looking at it, we can notice that many work 40 hours per week, while others work part-times and work more than 40 hours a week.  

```{r} 
par(mfrow=c(1,2))	
summary(hours)		
hist(hours) #histogram of working hours 
plot(hours, logwg)
```

### 1.1. Work hours as a linear measure 

Now, we will create Model 1, which sets the variable work hours as a linear measure. We will also include control variables, education and race. That is because, in the earlier assignments, we have observed that effects of education and race on earnings are statistically significant. I also looked at the potential of including the interacting variable, educ*race, which was proved to be statistically significant when looking at only those two variables. Surprisingly, however, when age and work hours variables were included, the effect of the interacting variable turned out to be not significant. Therefore, we will omit this interacting variable in Model 1 and the following models in this assignment, while keeping education and race as control variables. 

```{r}
model1<- lm(logwg~ age + hours + educ + race, data=cps) #linear measure of work hours
summary(model1)
```

According to Model 1, which treats both age and work hours as linear measures, all independent variables are statistically significant. Thus, changes in each independent variables are associated to changes in earnings. The adjusted R-squared, 0.28, indicates that Model 1 fits observations to a moderate extent. 



### 1.2. Work hours as categorical measures

As creating categorical independent variables can help us identify non-linear effects, we will recode work hours into categorical variables. Thus, we will create 3 work hours groups: part-time (less than 30 hours per week), full-time (between 30 to 40 hours per week), and over-tie (over 40 hours per week). 

```{r}
part<- as.numeric(hours < 30) #categorical measures of work hours
full <- as.numeric(hours >= 30 & hours <= 40) 
over <- as.numeric(hours > 40)
```


In order to create a linear model with categorical variable, we should set one group as the referent. We will create two models, Model 2 and 3, with different group as the referent in order to examine the possibility of non-linear effect. 

Model 2 sets full-time group as the referent and include educ, race, and educ*race as control variables. 

```{r}
model2 <- lm(logwg~ age+ part+ over + educ + race) #full time is the referent
summary(model2)
```

When the full-time group is set as the referent, the coefficients for the part-time and over-time groups are significantly different. Thus, the effect of work hours on earnings varies for different values of work hours. Moreover, the effect of work hours is negatively associated with earnings for the part-time group, as compared to the full-time group. 

In comparison, Model 3 below sets the part-time group as the referent. 

```{r}
model3 <- lm(logwg~ age + full + over + educ + race) #part time is the referent
summary(model3)
```

Similarly, when part-time group is set as referent, the coefficients for full-time and over-time groups are  significantly different. This reaffirms that the effect of work hours on earnings varies across the number of hours, and therefore, non-linear. 
In both Model 2 and 3, all independent variables are statistically significant. The adjusted R-square values for both models are similar. When compared to Model 1, Model 2 and Model 3 fit comparatively better, indicating the model improvement when we treat work hours variable as categorical instead of linear measures. 



## 2. Effect of age

Age is often used as a proxy for labor market experience, and it is commonly accepted that if one has more market experience, they would be likely earning more. However, the increase may not be linear, as some might experience more or less gains in average earning at a certain point of their career. Moreover, the wage may decrease past certain age as they approach retirement. Therefore, I believe that there is a theoretical reason to suspect that the relationship between age and earnings might be non-linear. We will investigate such possibility by using a quadratic term. 

```{r}
par(mfrow=c(1,2))
plot(age, logwg)

age.sq<-age^2	#create a new variable, age-squared
plot(age.sq, logwg)		
```


Model 4 below examines the effect of age in quadratic term with relevant control variables. 

```{r}
model4 <- lm(logwg~age + age.sq + hours + educ + race)
summary(model4)	
```

According to Model 4, all independent variables, including age in the squared term, are statistically significant. The signs of coefficients for linear age variable and quadratic age variable are different, suggesting that the effect of age on earnings change its direction at one point. This point can be found by a simple calculation below: 

```{r}
#  x = -B1/2*B2
age.change <- ((-1)*1.134e-01)/(2*-1.173e-03)
age.change
```

Thus, according to Model 4, the effect of age on earnings change its direction around age 48. 

Finally, we will determine whether Model 4, which includes age in quadratic term, fits better than Model 1, which specifies age as a linear measure, using incremental F-test. Looking at the test summary below, we can see that the F-ratio is very significant. Therefore, we can reject the null-hypothesis of linearity and conclude that age also has non-linear effects on earnings. 

```{r}
anova(model1, model4)  
```
