---
title: "Assignment 1: Bivariate Regression"
author: "Serena Choi"
date: "1/23/2022"
output: github_document
---

## Bivariate Regression and Residuals

In this assignment, we will perform a bivariate linear regression analysis of the OECD welfare state data. The objective is to test the null hypothesis that governments controlled by left/labor parties do *not* have more egalitarian welfare states. 

```{r}
oecd <- read.csv(file="oecd.csv")
```


### Research question 

This research question asks whether there is a relationship between welfare state development and the legislative & cabinet seat shares for left or labour parties among the OECD countries.

We can write a regression equation: 
*y = B0 + B1x + e* 
where y (response variable) is represented by **decom** and x (explanatory variable) is represented by **left**. 


Our null hypothesis is as follows: 
*H0 = B1 = 0*
which means that the slope of the regression equation is 0. 


### Estimating a model 

We will examine the relationship between gdp and decom.  
The *slope* of the regression equation, B1, can be found using the following equation: cov(x,y)/var(x)

```{r}
B1 <- cov(oecd$left, oecd$decom)/var(oecd$left)
B1
```

The intercept is as follows: B0 = ybar - B1 * xbar

```{r}
B0 <- mean(oecd$decom) - (B1*mean(oecd$left))
B0
```

Finally, the linear regression model and its summary statistiscs are as follows: 

```{r}
model1 <- lm(decom ~ left, data=oecd)
summary(model1)
```

The key analyses from the model summary table above are: 

1. The intercept, B0, is 22.6877. As we discussed in the class, this number does not have much meaning as countries rarely have 0 seat shares of left or labor parties in reality. 

2. The slope, B1, is 1.6140, indicating a positive relationship between the two variables (decom & left). It can be interpreted that the 1.6140 unit increase in y (welfare state development) for a one unit increase in x (left seat share). 

3. Both the t-statistic of 2.583 and the p-value of 0.02002 indicate that we can reject the null hypothesis. Therefore, the changes in left/labor party seats are related to changes in welfare development. 
4. However, the R-squared value is pretty low, indicating the model is not a good fit to data overall. 



### Residuals 

The scatterplot of data and the regression line look like the below: 

```{r}
plot(oecd$left, oecd$decom, main="OECD welfare state per left political party representation", 
     xlab="left", ylab="decom")
abline(a=B0, b=B1, col="blue")
text(oecd$left, oecd$decom, labels=oecd$X, cex=0.75, 
     font=1)
```


By examining the residuals, we can see that in countries such as Canada, Ireland, Italy and Finland, left party representation is a strong predictor of decommodification. On the other hand, In countries such as USA, Australia, NZ, Netherlands, UK, Denmark and Norway, left party representation is a poor predictor of welfare state development. 
