---
title: "Assignment 3: Categorical Variables and Interaction Effects"
author: "Serena Choi"
date: "2/6/2022"
output: github_document
---

For this assignment, we will use the U.S. CPS data. 
```{r}
cps <- read.csv(file="cps.csv")
race <- cps$race
age <- cps$age
educ <- cps$educ
logwg <- cps$logwg
```

## Model Estimation

### 1. Effect of age and education on weekly wages

#### Education as a continuous measure

**Model 1** considers age and education as both continuous variables. Setting log weekly wages as dependent variable, it examines the effect of education on earnings, net of age, and vice versa. 

```{r}
model1 <- lm(logwg ~ age + educ)
summary(model1)
```


#### Education as a categorical variable 

Instead of a continuous variable, education can be modeled as a categorical variable. We will create four categorical variables depending on education level: 

```{r}
lths <- as.numeric(cps$educ < 12)
hs <- as.numeric(cps$educ == 12)
somecoll <- as.numeric (cps$educ >12 & cps$educ < 16)
ugrad <- as.numeric(cps$educ >= 16)
```

When modelling with these categorical variables, we will set the first group (education less than high school; "lths") as referent. Hence, the second model looks like the following:

```{r}
model2 <- lm(logwg ~ age + hs + somecoll + ugrad)
summary(model2)
```

### 2. Interaction effect between education and race 

The simple additive model of age and race is as follows:

```{r}
model3 <- lm(logwg ~ educ + race)
summary(model3)
```

When we add the interaction between age and race, the model changes as the below:

```{r}
model4 <- lm(logwg ~ educ + race + educ*race)
summary(model4)
```

## Interpretation

**Model 1** below is based on the assumption that both age and education are *continuous*. That is, a unit increase in age is associated with, on average, a 0.022 unit increase in log weekly wage, net of education. Similarly, a unit increase in education is associated with, on average, 0.089 unit increase in log weekly wage. Looking at both t-values, compared to standard errors, and p-values, we can conclude that both variables are statistically significant, and therefore, we can reject the null hypotheses, where coefficients are 0. Looking at the adjusted R-square value, this model accounts for about 21% of variance of y. While it is not terrible, the model does not fit the observations well. 

```{r}
summary(model1)
```


**Model 2** still looks at the effect of age and education on weekly wages, however, the difference from Model 1 is that Model 2 considers education as a *categorical variable*. This allow us to compare more than two groups and examine effect of educational transitions on wages. In this model, less than high school group is set as the referent. Hence, the intercept, 4.770, represents the log weekly wages for high school dropouts, net of age. The coefficient estimates for the other variables show the average differences in log weekly wages depending on education levels. We can test the hypothesis that all the slope coefficients are 0 using a basic F-test. Looking at the F-statistic value, we can reject the null and determine that all variables here are statistically significant. The fit of Model 2 is similar to that of Model 1. 
```{r}
summary(model2)
```


As you can see in the plot below, Model 2 assumes that the slopes of each group are parallel. That is, the net effect of one variable is the same regardless of levels of the other variables in the model. 

```{r}
plot1 <- plot(age, logwg, main = "Effect of Age and Education Transition on Weekly Earnings")
abline(a=4.77, b=0.022, col="red")
abline(a=4.77+0.39, b=0.022, col="blue")
abline(a=4.77+0.39+0.5, b=0.022, col="green")
abline(a=4.77+0.39+0.5+0.84, b=0.022, col="orange")
legend(55,10, c("lths", "hs", "somcoll", "ugrad"), col=c("red", "blue", "green", "orange", pch(c("-"))))
```


**Model 3** investigates the idea that the effect of education on earnings may vary for black and white men in the U.S. Note that race here is a dummy variable. Looking at the summary statistics below, we can analyze that on average, white men earn 0.179 unit more in weekly log wages than black men. This difference is statistically significant. This model does not fit well as the adjusted R-squared value is low. 
```{r}
summary(model3)
```

Similar to Model 2, Model 3 also assumes that the slopes of both groups are parallel. The only difference is in the intercept, which indicates the average difference in earnings for black and white men. 

```{r}
plot2 <- plot(educ, logwg, main= "Effect of Education & Race on Earnings: Additive")
abline(a=4.68, b=0.09, col="red")
abline(a=4.68+0.18, b=0.09, col="blue")
legend(20,10, c("Black", "White"), col=c("red", "blue"))
```

<br>
Different from Model 3, **Model 4** considers an interaction effect between education and race by adding the variable, education*race. Given the summary statistics below, the race coefficient estimate, 0.004, is the difference between average log wages for black and white men in the U.S. This is much smaller than that of Model 3. However, this difference is not statistically significant when looking at t-value. The interaction variable, however, is statistically significant at 0.05 level. 

```{r}
summary(model4)
```
<br>
The major difference between Model 3 and Model 4 is that Model 4 does not assume parallel slopes. This becomes clear when we plot Model 4: 

```{r}
plot3 <- plot(educ, logwg, main= "Effect of Education & Race on Earnings: Interaction")
abline(a=4.84, b=0.08, col="red")
abline(a=4.84+0.00, b=0.08+0.01, col="blue")
legend(20,10, c("Black", "White"), col=c("red", "blue"))
```

The fact that the two groups (black and white) have different slopes indicates that the net effect of education on earnings may vary depending on race. We can compare Model 3 and Model 4 by conducting an incremental F-test. Looking at the ANOVA summary statistics below, the F ratio is 5. An F-ratio of 5 on 1 and 31635 df is significant at 0.05 level. Therefore, we can reject the null hypothesis of parallelism and conclude that there is an interaction between age and race. 

```{r}
anova(model3, model4)
((18120-18117)/(1))/(18117/31635)
```
